---
title: "Reproducible Data Peer Assessment 1"
author: "Sammy"
date: "4/13/2021"
output: md_document
keep_md: true 
---
Set global options
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

Load basic packages
```{r library, include=TRUE}
library(ggplot2)
library(dplyr)
library(lattice)
```

## Loading and preprocessing the data
1. Code for reading the dataset and/or processing the data.
```{r loaddata}
unzip("activity.zip")
data <- read.table("activity.csv", sep = ",", header = TRUE, na.strings = "NA", colClasses = c("numeric", "Date", "factor"))
```

Observe the dataset
```{r readdata}
head(data)
summary(data)
```

Tidy the data
```{r tidydata}
data1 <- na.omit(data)
data1$date <- as.Date(data1$date, format = "%Y-%m-%d")
str(data1)
```

## What is the mean total number of steps taken per day

Calculate the total number of steps taken per day (omit NA values).
```{r totalsteps}
totalsteps <- aggregate(steps ~ date, data1, sum)
print(totalsteps)
```

Create a histogram of a total number of steps taken each day.
```{r histplot}
histplot <- ggplot(totalsteps, aes(steps)) + 
          geom_histogram() +
          xlab("Number of Steps in a Day") + 
          ylab("Frequency") +
          ggtitle("Histogram of Total Number of Steps per Day")
print(histplot)
```

Calculate and report the mean and median of the total number of steps taken per day

```{r meansteps}
mean <- mean(totalsteps$steps, na.rm = TRUE) 
```

```{r mediansteps}
median <- median(totalsteps$steps, na.rm = TRUE)
```

The mean is `r mean` and the media is `r median`.

## What is the average daily activity pattern?

Create a data frame consisting of the average number of steps for each interval.
```{r intstep}
intstep <- aggregate(data1$steps, list(interval = as.numeric(as.character(data1$interval))), FUN = mean)
names(intstep)[2] <- "average_steps"
```

Plot the a time series plot the the dataframe.
```{r timeplot}
timeplot <- ggplot(intstep, aes(interval, average_steps)) +
          geom_line() +
          xlab("5-minute Intervals") + 
          ylab("Average Number of Steps") +
          ggtitle("Time Series Plot of Average Number of Steps Taken per Interval")
print(timeplot)
```

```{r maxintstep}
max <- intstep[which.max(intstep$average_steps),]
```

The `r max$interval`th interval has the maximum average step of `r max$average_steps`.

## Imputing missing values

```{r missdata}
miss <- sum(is.na(data))
```
There is a total of `r miss` missing observations in the data.

Create a data frame where the NA values are replaced with the mean values for that interval.
```{r filldata}
filldata <- data
for (i in 1:nrow(filldata)) {
          if(is.na(filldata$steps[i])) {
                    filldata$steps[i] <- intstep[which(filldata$interval[i] == intstep$interval), ]$average_steps
          }
}
       
```

Validate that all NA values have been replaced. 
```{r missdata2}
miss2 <- sum(is.na(filldata))
```
There is a total of `r miss2` missing observations in the data.


Calculate the total number of steps taken per day with replaced NA values.
```{r totalsteps2}
totalsteps2 <- aggregate(steps ~ date, filldata, sum)
print(totalsteps2)
```

Create a histogram with the updated data frame.
```{r histplot2}
histplot2 <- ggplot(totalsteps2, aes(steps)) + 
          geom_histogram() +
          xlab("Number of Steps in a Day") + 
          ylab("Frequency") +
          ggtitle("Histogram of Total Number of Steps per Day")
print(histplot2)
```
Calculate and report the mean and median of the total number of steps taken per day

```{r meansteps2}
mean2 <- mean(totalsteps2$steps, na.rm = TRUE) 
```

```{r mediansteps2}
median2 <- median(totalsteps2$steps, na.rm = TRUE)
```

The mean is `r mean2` and the media is `r median2`.

## Are there differences in activity patterns between weekdays and weekends?

Add weekday information to the data frame.
```{r weekday}
filldata$weekdays <- factor(format(filldata$date, "%A"))
levels(filldata$weekdays)
```

Split the weekday column into "Weekday" and "Weekend" lists. 
```{r week}
levels(filldata$weekdays) <- list(Weekday = c("Monday", "Tuesday", "Wednesday",
                                              "Thursday", "Friday"),
                                  Weekend = c("Saturday", "Sunday"))
levels(filldata$weekdays)
```

Create a data frame consisting of the average number of steps for each interval.
```{r intstep2}
intstep2 <- aggregate(filldata$steps, list(interval = as.numeric(as.character(filldata$interval)),
                                           weekdays = filldata$weekdays), FUN = mean)
names(intstep2)[3] <- "average_steps"
```

Create a panel plot containing a times series plot on the average number of steps take across weekdays and weekends. 
```{r timeplot2}
timeplot2 <- ggplot(intstep2, aes(interval, average_steps)) +
          geom_line() +
          facet_wrap(~ weekdays, nrow = 2, ncol = 1) +
          xlab("5-minute Intervals") + 
          ylab("Average Number of Steps") +
          ggtitle("Weekday vs Weekend Time Series Plot of Average Number of Steps")
print(timeplot2)
```


